How to Not write Virtual Machines for Dynamic Languages

### 2007-05-27 talk/dyla2007/dyla.tex
More reorganization, and finished another subsection.

The present paper proposes to approach the implementation of dynamic
languages from a meta-level: virtual machines for such languages should
not be written by hand, but generated automatically ``around'' a
description of the language in the form of an interpreter for it.  We
argue that this approach gives many of the benefits usually expected by
an implementer when he decides to target an existing object-oriented
virtual machine.  It also gives other benefits that we will describe --
mostly in term of flexibility.  But most importantly, it lets a
community write a single source implementation of the language, avoiding
the time-consuming task of keeping multiple ones in sync.  The single
source can be used to generate either custom VMs for C-like
environments, or interpreters running on top of OO VMs.  It makes it
practical to experiment with large changes to the language and with
entirely new languages, like domain-specific languages, while at any
time being able to run the implemented language in a variety of
environments, from C/Posix to the JVM to .NET.

@@

Our approach -- a single ``meta-written'' implementation -- naturally leads
to language implementations that have various advantages over the
``hand-written'' implementations.  First of all, it is a single-source
approach -- we explicitly seek to solve the problem of proliferation of
implementations.

Separating the implementation of a language in a high-level
``description'' and a custom translation framework has also many
advantages -- in our opinion significant enough to hint that
meta-programming, though not widely used in general-purpose programming,
is an essential tool in a language implementer's toolbox.

@@

@@

Of course, the price to pay is the need for a translation toolchain
capable of analyzing and transforming the high-level source code and
generating lower-level output in various languages.  Of course, the
translation toolchain, once written, can be reused to implement other
languages as well.  Even so, we found that the required effort that must
be put into the translation toolchain in the first place is still much
lower than that of writing a complete, commercial-quality OO VM.  A
reason appears to be that we could design our translation toolchain
specifically for our needs, i.e. a language implementer's needs, instead
of for general-purpose usage.

...

At the level of the translation framework, the ability to change or
insert new whole-program transformations makes some aspects of the
interpreter easier to deal with.  By ``aspect'' we mean, in the original
AOP sense, a feature that is added to an object program by a
meta-program.  The most obvious example in our context is the garbage
collector for the target environments that lack it.  The hand-written C
source of CPython is littered with macro calls that increment and
decrement reference counters.  Our translation framework can insert such
macro calls automatically -- in fact, we have a choice of GCs, and
reference counting is only one of them (not a particularly efficient
one, either).  Some GCs require different transformations of the code.
By contrast, supporting more than one GC in CPython is close to
impossible without forking the whole source code base.

In fact, there is a well-known example of a CPython fork: Stackless
Python \cite{stackless}, which adds support for coroutines.  This is very hard to
do in CPython because the interpreter is written in a highly recursive
style.  Stackless Python required large-scale changes; it is not merged
back into CPython due to the pervasive increase in complexity that it
requires.  In PyPy, though, an optional ``stackless'' transformation is
able to turn the Python interpreter -- also written in a simple highly
recursive style -- into an efficient variant of continuation-passing
style (CPS), enabling the usage of coroutines in the translated
interpreter.  For more details and other examples of translation-level
transformations, see \cite{D07.1}.

\subsection{Integration with a host OO VM}

A good example of this is to compare it against the task of hand-writing
an implementation for a specific, choosen OO VM.  The latter requires
not only good knowledge of the OO VM in question and its object model --
it requires the language implementer to fit the language into the
imposed models.  For example, it is natural to map classes and instances
of a dynamic object-oriented language to the OO VM's notion of classes
and instances, but this might not be a simple task at all if the models
are substantially different and/or if the OO VM is essentially less
dynamic than the language to implement.  In our approach, this task
is done at two levels: in a first step, while writing the interpreter,
the implementer does not need to worry about integration with an OO VM's
object model.  Of course, the integration effort does not simply vanish
-- indeed, a simple translation of such an interpreter to a given OO VM
would give an interpreter for the dynamic language which is unable to
communicate with the host VM (which might already be interesting in
specific cases, but not in general).  Integration comes as a second
step, and occurs at a different level, by introducing mappings between
the relevant classes of the interpreter and the corresponding classes of
the OO VM.  As of yet we have no evidence that this makes the total
integration effort much lower or higher; the point is that it has proven
possible to take the single, OO VM-independent source code of PyPy's
Python interpreter, and produce from it a version that runs in and
integrates with the rest of the .NET environment -- by writing
\emph{orthogonal} code.

\subsection{Getting good GCs and tools is possible}
XXX

>>>>

The present paper proposes to approach the implementation of dynamic
languages from a meta-level: virtual machines for such languages should
not be written by hand, but generated automatically ``around'' an
interpreter playing the role of a high-level description of the language.  We
argue that this approach gives many of the benefits usually expected by
an implementer when he decides to target an existing object-oriented
virtual machine.  It also gives other benefits that we will describe --
mostly in term of flexibility.  But most importantly, it lets a
community write a single source implementation of the language, avoiding
the time-consuming task of keeping multiple ones in sync.  The single
source can be used to generate either custom VMs for C-like
environments, or interpreters running on top of OO VMs.  It makes it
practical to experiment with large changes to the language and with
entirely new languages, like domain-specific languages, while at any
time being able to run the implemented language in a variety of
environments, from C/Posix to the JVM to .NET.

@@

Our approach -- a single ``meta-written'' implementation -- naturally
leads to language implementations that have various advantages over the
``hand-written'' implementations.  First of all, it is a single-source
approach -- we explicitly seek to solve the problem of proliferation of
implementations.  In the sequel, we will show that this goal can be
achieved without giving up on the advantages of hand-written
implementations for OO VMs.  Moreover, there are additional advantages
-- in our opinion significant enough to hint that meta-programming,
though not widely used in general-purpose programming, is an essential
tool in a language implementer's toolbox.

@@

\subsection{Separation of concerns}

At the level of the translation framework, the ability to change or
insert new whole-program transformations makes some aspects of the
interpreter easier to deal with.  By ``aspect'' we mean, in the original
AOP sense, a feature that is added to an object program by a
meta-program.  The most obvious example in our context is the insertion
of a garbage collector (chosen among several available ones) for the
target environments that lack it.  Another example is the translation of
the interpreter into a form of continuation-passing style (CPS), which
allows the translated interpreter to provide coroutines even though its
source is written in a simple highly recursive style.  For more details
and other examples of translation-level transformations, see
\cite{D07.1}.

A more subtle example of separation of concerns is the way our generated
implementations can be integrated with a host OO VM.  As mentioned
above, an implementer deciding to directly target a specific OO VM needs
not only good knowledge of the OO VM in question and its object model --
he must fit the language into the imposed models.  Instead, in our
approach this task is done at two levels: in a first step, a stand-alone
interpreter is written -- which, if translated to a given OO VM, would
simply give an interpreter for the dynamic language which is unable to
communicate with the host VM.  Integration comes as a second step, and
occurs at a different level, by introducing mappings between the
relevant classes of the interpreter and the corresponding classes of the
OO VM.

@@

What are the efforts required to develop a translation toolchain capable
of analyzing and transforming the high-level source code and generating
lower-level output in various languages?

Although it is able to generate, among other things, a complete custom
VM for C-like environments, we found that the required effort that must
be put into the translation toolchain was still much lower than that of
writing a good-quality OO VM.  A reason seems to be that we could design
our translation toolchain specifically for our needs, i.e. a language
implementer's needs, instead of for general-purpose usage.  Of course,
the translation toolchain, once written, can also be reused to implement
other languages, and possibly tailored on a case-by-case basis to fit
each implementer's need.  The process is incremental: we can add more
features as needed instead of starting from a maximal up-front design,
and gradually improve the quality of the tools, the garbage collectors,
the various optimizations, etc.

Let us expand on the topic of the garbage collector, which for C-like
envrionments is inserted into the generated VM by a transformation step.
We started by ignoring the issue and just using the conservative Boehm
\cite{Boehm} collector for C.  Later, we experimented with a range of
simple custom collectors - reference counting, mark-and-sweep, etc.
Ultimately, though, more advanced GCs will be needed to get the best
performance.  It seems that RPython, enhanced with support for direct
address manipulations, is a good language for writing GCs, so it would
be possible for a GC expert to write one for our translation framework.
However, this is not the only way to obtain good GCs: we will soon
investigate a more practical course of action, which is to reuse
existing GCs.  A good candidate is the GCs written in the Jikes RVM
\cite{JikesGC}.  As they are in Java, it should be relatively
straightforward to add a translation step that turns one of them into
RPython (or directly our RPython-level intermediate representation) and
integrate it with the rest of the program being translated.

In summary, developing a meta-programming translation toolchain requires
some work, but it can be done incrementally, it can reuse existing code,
and it gives a toolchain that is itself highly reusable and flexible in
nature.
